---
title: "wrangling-and-cleaning"
format: html
editor: visual
---

## Load Libraries
```{r}
library(dplyr)
library(tidyr)
library(readr)
```


## Running Code
bit.ly/data-wrangle-lesson-data
```{r}
catch_original <-read.csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1")
```

### Initial Data Exploration
```{r}
colnames(catch_original)
```


```{r}
unique(catch_original$Region)
```


```{r}
summary(catch_original)
```
## Pipe Operator
catch_data <- catch_original %>% select("Reg)
### Select Columns 
```{r}
catch_data <-catch_original %>% select(Region,Year,Chinook,Sockeye,Coho,Pink,Chum)
```
## Quality Check
```{r}
dplyr::glimpse(catch_data)
```

### Change column content

GOAL: convert the Chinook catch values to "numeric" and overwrite the old Chinook column
```{r}
catch_clean<-catch_data%>%
  mutate(Chinook=as.numeric(Chinook))
```
```{r}
dplyr::glimpse(catch_clean)
```
##Find which rows has NA
```{r}
which(is.na(catch_clean$Chinook))

```
##Use pipe find row instead
```{r}
catch_clean$Chinook %>%is.na()%>%which()
```
GOAL: change the value of the Chinook column to 1 if the value is equal to the letter I
```{r}
catch_data <-catch_data %>%
  mutate(Chinook = if_else(condition = Chinook == "I", true = "1", false = Chinook), Chinook = as.numeric(Chinook))
head(catch_clean)
```

```{r}

catch_long <-catch_clean%>%pivot_longer(cols= -c(Region,Year), 
                           names_to = "species", 
                           values_to="catch")
```

```{r}
catch_wide <- catch_long %>%
    pivot_wider(names_from = species,
                values_from = catch)

head(catch_wide)
```


```{r}
catch_long <- catch_long %>%
    rename(catch_thousands = catch)

head(catch_long)
```
```{r}
catch_long <- catch_long %>%
    mutate(catch = catch_thousands * 1000)

head(catch_long)
```

```{r}
catch_long <- catch_long %>%
    mutate(catch = catch_thousands * 1000) %>%
    select(-catch_thousands)
```

##Summarize
```{r}
mean_region <- catch_long%>%
group_by(Region)%>%
  summarize(mean_catch=mean(catch))
```
```{r}
sse_catch <- catch_long %>%
    filter(Region == "SSE") %>%
    filter(species == "Chinook")

head(sse_catch)
```

```{r}
mil <- catch_long %>%
  filter(catch>1000000)
head(mil)
```

##filtering

```{r}
sites_df <- data.frame(site = c("HAW-101",
                            "HAW-103",
                            "OAH-320",
                            "OAH-219",
                            "MAI-039"))

df <-sites_df %>%
    separate(site, c("island", "site_number"), "-")
```
```{r}
df %>%
  unite(code, island, site_number, sep = "-")
```


#Sort Data
GOAL: recalculate mean catch by region, and then sort the output by mean catch
```{r}
mean_region <- catch_long %>%
    group_by(Region) %>%
    summarize(mean_catch = mean(catch)) %>%
    arrange(desc(mean_catch))

head(mean_region)
```

